Formal Grammars as Types in Non-commutative Linear-Non-Linear Type Theory

### 2024-07-07 POPL/paper.tex
scaffolding the type theory section

\section{Kleene Category}
\steven{No changes pasted from LICS draft}
\steven{Are these now ``Chomsky categories''? or is that disinct?}

@@

\begin{definition}
  A model is a locally Cartesian category with two hierarchies of
  universes $\{L_i\}_{i\in \nat}$ and $\{U_i\}_{i\in \nat}$ such that
  every $L_i$ and $U_i$ are $U_{i+1}$-small. Furthermore, we require
  $U_i$ to be closed under dependent products and sum,
  $L_i$ to be closed under the Kleene category connectives,
  dependent products, left and right closed structures, with
  a type constructor $G : L_i \to U_i$ and a linear dependent sum
  going the other direction.
\end{definition}

@@

@@

%% Vaughan Pratt on the residual operator

\section{Rest of paper}
\steven{Everything below still needs to be edited from my prelim paper}
\steven{Need to grab the canonicity proof from the LICS draft}

Substructural logics, like linear and separation logic, offer precise control
over how values from a type may be used, in particular how many times a value
may be used in a derivation. This control enables \emph{resource-sensitive}
reasoning, where we may treat a resource as \emph{consumed} after usage. In the
context of parsing, we may treat the characters composing a string as finite resources
that are consumed at parse-time.

Dependent type theory extends the simply-typed $\lambda$-calculus by allowing type
definitions to contain the occurrence of a term. In lieu of set-based
foundations, this extension can then be taken as a foundation for mathematics.
The expressive power unlocked by this insight has enabled the development of
dependently-typed proof assistants
such as Coq or Agda. Within each of these languages, systems may be given very
precise functional specifications of their behavior.

In 1995, Benton et al.
\cite{bentonMixedLinearNonlinear1995} first gave categorical models of a logic
that synthesized linear and non-linear logic. Later, Krishnaswami et al.
\cite{krishnaswami_integrating_2015} built upon this work to develop a type
system that unifies both linear and
dependent types --- referred to as a \emph{linear-non-linear} (LNL) type theory. In
such settings, we have two sub-theories --- one linear and one non-linear. The
linear sub-theory embeds ordinary linear types without dependence, while the
non-linear sub-theory embeds standard Martin-L\"of type theory. The interplay of
linear types with dependence lives at the boundary between these two fragments.
Benton et al.\ gives an accounting of how we may jump between these
fragments via a monoidal adjunction $F \dashv G$ between linear terms and
non-linear terms.

To the end of defining real verified
parsers, in our LNL theory we define a parser as a term,

\crefname{equation}{}{}
\begin{equation}
  \label{eq:parser}
  w : \String \vdash \mathsf{parse}_{g}(w) : g \oplus \top
\end{equation}

We will investigate the precise meaning of judgments like \cref{eq:parser} in
\cref{sec:tt}, but first let's get an intuitive feel for the meaning behind such a
term. On the left of the turnstile occurs a single string $w$, we may think of
this describing what resources are linearly available to us. Using each piece of
our the context on the left of the turnstile, we build a term
$\mathsf{parse}_{g}(w)$ of type $g \oplus \top$. Notably, the type being
instantiated is a sum type, and the data of which side we inject into denotes
whether the string is accepted or rejected by the parser. That is all to say, a
parser takes in an input string and either matches it to the desired grammar or
rejects the string.

Practical parsers are often implemented via an abstract automaton. For instance,
a parser may translate a grammar into a state machine recognizes the
same language as the original grammar, as the automata
point-of-view lends itself to straightforward execution. The
strength of our syntax is that it allows us to describe
these automata as grammars themselves. Moreover, the
equivalence between the grammar and automata presentations
is derivable as a proof term within the theory. This allows
us access to simultaneously internalization of the equivalence
between grammars and automata as well as the ability to
export to automaton implementations of a parser that are executable.

To realize an executable implementation of our type theory,
we shallowly embed it in Agda. Because our implementation is
defined as a function in Agda, we can leverage the Agda
ecosystem for code generation. Further, in lieu of
implementing a complete language from scratch, we make use of the
Agda proof system as a setting to guarantee correctness of
our generate parsers.

Within this paper, we give our progress in defining a verified
parser for regular expressions. The construction of such a parser entails
the definition of a parser for deterministic finite automata (DFA) grammars,
and an equivalence proof between DFAs and regular expressions. We have thus far
completed the construction of a parser for DFAs $D$,

\[
  w : \String \vdash \mathsf{parse}_{D}(w) : D \oplus \top
\]

and half of the equivalence proof between DFAs and regular expressions, both
formalized in Agda. Upon completion of the second half of this proof, we would
have a system where a user could simply give a regular grammar and then extract
an intrinsically verified parser for it. In future work, we
hope to show an analogous construction involving pushdown
automata (PDAs) and context-free grammars.

\section{A Syntax-Independent Notion of Syntax}
\label{sec:synindsyn}

% \max{this section doesn't flow naturally from the intro, which is all about the
%   type theory and parsing. This is where you want to outline in the intro and
%   possibly say here that we want to revisit ``formal grammar theory'' and then
%   we will make a type theory that has formal grammar theory as its semantics.}

In this section we will revisit some foundational ideas in formal
grammar theory, before giving a type
theory for which formal grammars serves as its semantics.

Formal languages and grammars are among some of the oldest and most studied areas in
theoretical computer science. The developments of the 1950s and 1960s gave rise
to expressive formalism for reasoning about grammars, and further practical
algorithms for constructing parsers for regular and context-free grammars.

A grammar $G$ is often described generatively as a set of production rules for
generating strings. Often this is presented as a finite set of nonterminal
symbols $N$, a finite set of terminal symbols $\Sigma$, and a finite set of rewriting
rules for substitution of nonterminals. In this setting, we continually apply
rewrite rules until we are left only with a string of terminals
$w \in \String$. We will give a more precise notion of
grammar in a moment, but for now let us take this intuitive view of grammars as
string generators.

One of the primary objects of study is that of a \emph{formal language} over
a fixed alphabet $\Sigma$. A language $L$ is a \emph{subset} of strings
$L \subseteq \String$. This notion of formal language gives an intuitive
set-based semantics for grammars. That is, we may interpret a grammar $G$ as the
language $L(G)$ that is generated by it. We are yet to give any precise
definition of what we mean by ``grammar'', although we note that any suitable
notion of grammar can be given a language semantics as above.

Formal languages serve as a basis for mathematically specifying a
\emph{recognizer} --- a
function $r : \String \to \Bool$ that determines if a string belongs to a
given language. However, they are insufficient for specifying a \emph{parser}
--- a function that decomposes a string into semantic components that adhere to
the structure of a formal grammar. In practice, parsers don't just tell you
if a string is valid or invalid. Instead, they emit structured data such as
abstract syntax trees (ASTs).

This motivates our first conceptual contribution, a simple and
syntax-independent notion of formal grammar:

We say that a grammar $G$ associates to every string $w$ the set\footnote{We
  make little comment as to which foundations we use to encode sets. We argue\max{drop ``We argue''}
  that our construction is agnostic to the particular foundation, and so without
  loss of generality\max{incorrect use of ``without loss of generality''} we may take a class of small sets, a Grothendieck universe,
  a type theoretic universe, or any similar foundation. Here we choose a type
  theoretic universe.} $G w$ of parse
trees showing that $w$ matches $G$.

The simplest class of formal grammar on the Chomsky\max{first mention of Chomsky hierarchy, why not bring it up earlier when you are describing what a grammar is? Also cite}
hierarchy are the \emph{regular expressions}

\max{inaccurate reflection of common terminology. These are regular expressions,
  regular grammars usually means a CFG with a restriction on recursion that
  ensures you only allow Kleene star.}.

\begin{gather*}
  \mathsf{Char} \in \Sigma \\
  \mathsf{Reg} ::= \mathsf{Char}~|~I~|~\mathsf{Reg} \otimes \mathsf{Reg}~|~\mathsf{Reg} \oplus \mathsf{Reg}~|~\mathsf{Reg}^{*}
\end{gather*}

Here, we use a non-standard syntax inspired by linear logic. Disjunction of
regular expressions is written $\oplus$, and concatenation of regular
expressions is written like the linear tensor product $\otimes$. The
empty grammar is sometimes written in the same manner as the empty string,
$\varepsilon$, but we instead choose to represent the empty grammar by $I$. This
is also linear logic syntax to represent the unit $I$ of the operation $\otimes$.

To see an illustration of the notion of grammar given in \cref{def:grammar}, consider the regular expression
$G = (a \otimes b)^{*} \otimes a$ --- a sequence of 0 or more $ab$'s followed by
an $a$. We may think of $G$ as a function that takes in strings and outputs all
of the parse trees of that string. Passing the input string $baa$ through $G$
would output the empty set, because there are no $G$-parses of $baa$. However,
passing $aba$ through $G$ would output a singleton set, as there is only a
single $G$-parse of $aba$. Namely, the one that breaks up $aba$ as $(ab)a$.

Observe that for any input string $w$ for the above grammar $G$, $G(w)$ is
either empty or a singleton set. Either the string isn't parsed by $G$,
or it is parsed by $G$ for exactly one reason. That is, $G$ is
\emph{unambiguous}. If we instead consider the grammar
$H = (a \otimes b)^{*} \oplus (b \otimes a)^{*}$ there are now ambiguities. For
instance, what is the action of $H$ on the empty string $\varepsilon$?
The empty string $\varepsilon$ matches both $(a \otimes b)^{*}$ and $(b \otimes a)^{*}$, so
$\varepsilon$ can match either the left or the right pattern
of $H$. As a result, $H(\varepsilon)$ is a two-element set.

\[
  L_{G} = \{ w \in \String : G w \text{ is inhabited} \}
\]

As they are just sets, formal languages naturally are
endowed with a partial
order via the subset inclusion ordering. Similarly, formal grammars are
naturally endowed with the structure of a \emph{category} where a morphism
$f : G \to H$ is given by a \emph{family} of functions:

\[
  f^{w} : Gw \to Hw, w \in \String
\]

Intuitively, we read a morphism of grammars as a \emph{parse transformer}: $f$
translates a
$G$-parse of $w$ into an $H$-parse of $w$. The notion of isomorphism induced by
this categorical structure encodes that two grammars are equivalent if there is
a bijective translation of parses. Isomorphisms in the category of
grammars are precisely Chomsky's notion of \emph{strong equivalence} of
grammars \cite{chom1963}.

The category of grammars is quite rich in structure. It can
equivalently be described as the functor category $\Set^{\String}$, where
$\String$ is a discrete category. That is, the objects in $\String$ are strings
and the only morphisms are identity morphisms. Such functor categories into
$\Set$ --- often called presheaf categories --- carry a remarkable amount of
structure. In fact, any presheaf category naturally models dependent type
theory, which we exploit to motivate the design of the formalism
presented in \cref{sec:tt}.

\section{A Type Theory to Reason About Formal Grammars}
\label{sec:tt}

@@

In this section we will walk through most of the syntax of our type theory. We
include the full syntax in \cref{sec:infer}.

@@

\begin{theorem}[Thompson]
  \label{thm:thompson}
  For $g$ a regular grammar $g$, there is an NFA $N$ that recognizes the same
  language as $g$.
\end{theorem}

>>>>

\section{Categorification of Formal Grammars}
\steven{Idk if this section has the right title, but I wanted to change it if we are adding ``chomsky categories''}

@@

\begin{definition}
  A \emph{Chomsky Category} is a locally Cartesian category with two hierarchies of
  universes $\{L_i\}_{i\in \nat}$ and $\{U_i\}_{i\in \nat}$ such that
  every $L_i$ and $U_i$ are $U_{i+1}$-small. Furthermore, we require
  $U_i$ to be closed under dependent products and sum,
  $L_i$ to be closed under the Kleene category connectives,
  dependent products, left and right closed structures, with
  a type constructor $G : L_i \to U_i$ and a linear dependent sum
  going the other direction.
\end{definition}

@@

\steven{TODO: move to related work section at the end}

@@

%% Vaughan Pratt on the residual operator
%%

\section{Non-commutative Linear-non-linear Type Theory as a Syntax for Formal Grammars}
\label{sec:tt}

In 1995, Benton et al.
\cite{bentonMixedLinearNonlinear1995} first gave categorical models of a logic
that synthesized linear and non-linear logic. Substructural logics, like linear and separation logic, offer precise control
over how values from a type may be used, in particular how many times a value
may be used in a derivation. This control enables \emph{resource-sensitive}
reasoning, where we may treat a resource as \emph{consumed} after usage. In the
context of parsing, we may treat the characters composing a string as finite resources
that are consumed at parse-time.

Building off of Benton et al., Krishnaswami et al.
\cite{krishnaswami_integrating_2015} developed a type
system that unifies both linear and
dependent types --- referred to as a \emph{linear-non-linear} (LNL) type theory. In
such settings, there are two sub-theories --- one linear and one non-linear. The
linear sub-theory embeds ordinary linear types without dependence, while the
non-linear sub-theory embeds standard Martin-L\"of type theory. The interplay of
linear types with dependence lives at the boundary between these two fragments.
Benton et al.\ gives an accounting of how we may jump between these
fragments via a monoidal adjunction $F \dashv G$ between linear terms and
non-linear terms.

\steven{Next paragraph is paper scaffolding for what I want to say and is total word salad rn}
In this paper, we instantiate an LNL theory to axiomatize the notion of a Chomsky category, and because $\Set^{\String}$ forms a Chomsky category, using the type theory corresponds to interesting formal grammar proofs. Internalization of previous results + use to build correct-by-construction parsers.

\steven{Expand here on how linearity has an intuitive parsing intrerpration, however working only in a simply typed system is not sufficient. i.e. dependence offers the right view on indexed grammars. }

@@

In this section we will walk through most of the syntax of our type theory. We
include the full syntax in \cref{sec:infer}. \steven{full syntax in appendix}

@@

\begin{theorem}[Thompson]
  \label{thm:thompson}
  For $g$ a regular grammar $g$, there is an NFA $N$ that recognizes the same
  language as $g$.
\end{theorem}
\steven{In Agda as a weak equivalence right now}
